Search techniques for gaussian mixture modelling with Minimum Message Length, with applications to the origin of our Galaxy

### 2018-10-30 article/ms.tex
words

\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have 
derived the bounds
%\begin{eqnarray}
%\sum_{k=1}^{K}\log\weight_k & \geq & -K\log{K} \nonumber \\
%\sum_{k=1}^{K}\log\weight_k & \leq & 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \nonumber
%\end{eqnarray}
\begin{equation}
-K\log{K} \quad \geq \quad \sum_{k=1}^{K}\log\weight_k \quad \geq \quad (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
\end{equation}
\noindent{}which  describe the smallest and largest contribution that the weights can contribute to the message length for any untrialled mixture with $K$ components. When making predictions for future mixtures, we record the optimal weights for each trialled mixture and calculate a \emph{uniformity fraction} $f_u \in [0, 1]$
\begin{equation}
f_u = \frac{\alpha - \sum\log\weight_k}{\alpha + K\log{K}}
\end{equation}
\noindent{}where for brevity $\alpha = (K-1)\log{2} + \log(N-2K+2)-K\log{N}$,
and $f_u$ the uniformity of the mixture from Equation \ref{eq:slw_nonuniform} ($f_u = 0$) to Equation \ref{eq:slw_uniform} ($f_u = 1$). We use estimates of $f_u$ from previously trialled mixtures to make predictions for the uniformity of future,
untrialled mixtures, and the message length required to encode the weights.
Given the theoretical bounds in Equations \ref{eq:slw_uniform} and \ref{ew:slw_nonuniform},
we clip the predictions of the expectation value on the message length required to encode the weights by these bounds.

@@

@@

@@

% TODO: Predicting the negative log-likelihood.

% TODO: Predicting the sum of the log of the determinant of the covariance matrices.

% Combining it together.

@@

We evaluate the message length of each component by rearranging Equation \ref{eq:objective_function} and multiplying the negative log-likelihood
$-\log{\mathcal{L}}(\vec\data|\vectheta)$ by the $N \times K$ responsibility matrix $\mathcal{R}$ 
%\begin{eqnarray}
%	\mathcal{R}_{nk} = \frac{\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}{\sum_{k=1}^{K}\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}
%\end{eqnarray}
such that the message length of the $k$th component is
\begin{eqnarray}
	I_k &=& -\sum_{n=1}^{N}\mathcal{R}_{nk}\log{\weight_{k}f(\data_{n}|\vecmean_k,\veccov_k)} -\frac{(D + 2)}{2}\log{|\cov_k|} \nonumber \\
		&& \cdots\ + \left(\frac{D(D+3)}{4}-\frac{1}{2}\right)\log{\weight_k} + \left(1 - \frac{D}{2}\right)\log{2} \nonumber \\
		&& \cdots + \frac{1}{K}\left[\log\Gamma(K) + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \frac{Q}{2}\log{N}\right]
\end{eqnarray}
\noindent{}and the total message length remains the sum of the message lengths of the components
\begin{eqnarray}
	I(\vec\data|\vec\psi, N, D) = \sum_{k=1}^{K}I_{k} \quad .
\end{eqnarray}

@@

When splitting the component with the largest message length, we split this component into two child components. Initially we 
set the means of the child mixtures as positive and negative eigenvalues along the principal
eigenvector of the component covariance matrix \citep{Kasarapu|Allison:2015}.
A matrix of child responsibilities is then
calculated based on the Euclidean distance between each datum and the child means (i.e., hard
association), which we use to estimate the covariance matrices and weights of the child
components. After the parameters of the child components are initialised, we run \emph{partial}
expectation-maximization to iteratively update the parameter estimates of the child components. 
By \emph{partial} expectation-maximization we refer to the process of keeping the parameters of
all other components fixed, and only updating the estimates of the parameters of the child
components \citep{Kasarapu|Allison:2015}. After partial E-M has converged, we integrate the child components with the rest of the mixture and run full E-M before the \textsc{Split} operation (line \ref{line:mbm-split} of Algorithm~\ref{alg:MBM}) is considered complete.

One advantage to the Message-Breaking Method is that the total number of E-M steps is relatively few because all partial E-M steps are performed from a near-optimal initialisation point. One possible disadvantage to the Message-Breaking Method is that there may be mixtures where the \textsc{Split} operation increases the message length by $\epsilon$, but a subsequent \textsc{Split} operation would \emph{decrease} the total message length by $>\epsilon$. As a result, the algorithm would converge to a local minima, not the global one. In the code implementation that we release with this work, we have implemented a \emph{Strict Message-Breaking Method} variant that allows the user to specify the number of successive $K$ trials that increased the message length, before adopting the mixture with the shortest message length. By default the Strict Message-Breaking Method requires five successive mixtures where the message length did not improve.

@@

The \emph{Message-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of computations. In the \emph{Message-Breaking Method} we
try to accomplish this using partial E-M to minimise the total number of computations
required to converge on a $K+1$ component mixture from a $K$ component mixture.
However, in the \emph{Message-Jumping Method} we seek to minimise the total number
of computations by optimizing a fewer number of mixtures. Instead of iteratively
perturbing a mixture from say $K = 10$ to $K = 30$, we use predictions of the
message length of untrialled mixtures and immediately `jump' to a component (
e.g., $K = 27$) that will provide the most improvement in our predictions of the
minimum message length over \emph{all} components. In doing so we seek a global 
message length minima with a fewer number of total E-M operations.

We describe the \emph{Message-Jumping Method} in Algorithm~\ref{alg:mbm}. In this
method we seek to predict the minimum message length for any number of $K$ components,
and then converge to the optimal mixture as quickly as possible. For this reason,
we initialise many mixtures that are evenly separated in $\log_{10}$ space from
$K = 1$ to $K = N/2$ in order to
quickly describe the minimum message length across all $K$. Specifically we
introduce a search hyperparameter $K_{inits}$ (defaults to 10) that sets the
number of mixtures to initialise -- and those initial $K$ values. 
We initialise each mixture using the K-means++ algorithm\cite{kmeanspp} and
optimize the mixture using EM as described earlier. We trial each of the $K_{init}$
mixtures in rank order of increasing $K$. If one of these mixtures fails to initialise
or optimise, this is likely because we could not perform Choleksy decomposition
of one (or more) of the covariance matrices. In practice this is typically because
the estimate of $K$ is much higher than $K_{true}$, leading to near-zero values
along the diagonal of the covariance matrix. In this situation we do not attempt
to initialise any more of the $K_{init}$ mixtures -- although higher $K$ values
can be trialled later -- and we being the search method.

The motivation for this initialisation process is to mimic the search strategy
that a human might take: What is an order-of-magnitude estimate on the number of
$K$? Or what is a ballpark number on the maximum number of $K$ that we should
trial? After we have that number, and an estimate on the message length for a
few sparsely-sampled numbers of components, we can make reasonable predictions
on the global minimum.

We record the message length for the optimal mixture at each trialled $K$, and
the model parameters for that mixture. With these quantities we predict the
message length of untrialled mixtures for every $K$ value from $K = 1$ to $K = \ceil{1.1\max(K_{trialled})}$ -- e.g., up to 10\% ahead of the largest previously trialled
$K$ value. The message length predicted from the optimised Gaussian processes
provide an expectation value on the message length for an untrialled mixture, as
well as a variance in that prediction (Figure \ref{fig:search-predictions}).
Before selecting the next value of $K$ to trial, we use the expectation value and
associated variance to compute the acquisition function (or the "expected
improvement") at each value of $K$ \cite{Jones:1998}.

We select the next $K$ value by choosing the value that will maximise the expected
improvement \todo{but this forces us out to large K, so ....}

\todo{initialisation with kmeans++, if it fails,....}

@@

\caption{The Component-Jumping Method.}
\label{alg:component-jumping-method}
{\bf Input:} $\vec\data$, $K_{i}$, $I_\epsilon$ \\
{\bf Output:} $\vec\psi$ \Comment{The optimal mixture parameters.}
\begin{algorithmic}[1]
\State s $\leftarrow (K_{i})^{-1}\log_{10}{N}$ \Comment{Initial step size in $K$.}
\State v $\leftarrow [0, 1, ..., K_{i}]$ 
\State $K_{inits} \leftarrow$ \Call{Power}{10, vs} \Comment{Set initial $K$ values to trial.}
\State list $K_{t}$ \Comment{Create a list to save trialled $K$ values.}
\For{$k\in K_{inits}$} 
  \State $K_{t}$.append($k$) \Comment{Remember that we trialled this $K$ value.}
  \try 
	  \State mix $\leftarrow$ \Call{KMeans$^{++}$}{k} \Comment{Initialise with K-means++.}
	  \State mix $\leftarrow$ \Call{ExpectationMaximization}{mix}
  \catch
  	  \State {\bf \sc Break} \Comment{Do not trial any higher $K$ values.}
  \finally 
      \State mixture[k] $\leftarrow$ mix 
      \State state[k] $\leftarrow$ \Call{MessageLength}{mixture[k]}
  \endtry
\EndFor
\State converged $\leftarrow$ FALSE
\While{not converged}
	\State $K_{predict} \leftarrow [1, \cdots, 2\Call{Max}{K_{t}}]$ \Comment{Predict future message lengths.}
	\State $I_{predict} \leftarrow$ \Call{PredictMessageLength}{$K_{predict}$}
	% argsort.
	\State $K_{next} \leftarrow$ \Call{ArgSort}{$I_{predict}$} \Comment{Sort $K$ by predicted message length.}
	\For{$k\in K_{next}$}
		% Check if the mixture at K was good?
		\If{$k\in K_{t}$}
			\State converged $\leftarrow$ TRUE
			\State {\bf \sc Break}
		\Else
			\State $K_{t}$.append($k$) \Comment{Remember that we trialled this $K$ value.}
	   	  \try 
		    \State mix $\leftarrow$ \Call{KMeans$^{++}$}{k} \Comment{Initialise with K-means++.}
	   	    \State mix $\leftarrow$ \Call{ExpectationMaximization}{mix}
		  \catch
  	        \State {\bf \sc Continue} \Comment{Move on to next $K$ trial.}
		  \endtry 
		\EndIf
	\EndFor
\EndWhile
\end{algorithmic}
\end{algorithm}

@@

We set the data dimensionality $D = 2$ after previous experiments found that the data dimensionality had only a small effect on the differential cost on the search strategies tested. We chose $N$ to vary uniformly in log space from $N = 10$ data points to $N = 10^{4.5}$. For each $N$ we set $K = N/10$, and then used these values of $N$, $D$, and $K$ to generate a synthetic dataset. The weights for each mixture were drawn from a Dirichlet distribution with concentration $\alpha = 100$ while ensuring that $\weight_k \in \left(\frac{2}{N}, 1-\frac{2(K-1)}{N}\right)$. The means for each component were drawn from $\mathcal{U}\left(-10, +10\right)$. The correlation coefficients  for the component covariance matrices were drawn from
\begin{equation}
	\rho \sim \frac{\left[1-\mathcal{U}(-1, +1)^2\right]^{\frac{1}{2}(s - 4)}}{\mathcal{B}\left(\frac{1}{2}, \frac{1}{2}(s - 4)\right)}
\end{equation}
\noindent{}where $s$ is an isotropy hyperparameter that describes the probability of correlations between data dimensions, which we fix as $s=100$. The diagonal variances for the component covariance matrices were drawn from $|\mathcal{N}\left(0, 0.1\right)|$. Given these randomly drawn parameters we generated $N\weight_k$ members for each component, rounding as necessary to ensure the total number of generated data points is kept at $N$.

@@

We use detailed chemical abundances from the Apache Point Observatory Galactic Evolution Experiment (APOGEE), an experiment run by the Sloan Digital Sky Survey (SDSS) collaboration. We make use of SDSS Data Release 12, which includes high-resolution, high S/N infrared spectra for $\sim$163,000 stars collected by APOGEE. Each observation
is an array of 8,575 pixels that have an associated rest-frame wavelength, and a
measurement of the stellar flux that is normalised relative to the spectral energy
distribution such that a flux value of 1 indicates no stellar absorption and a
flux value near 0 implies strong stellar absorption. All else being equal, a stronger
amount of absorption (lower normalised flux value) at a particular wavelength 
indicates that the star contains more of the chemical element that absorbs and
emits light at that wavelength. For example, sodium emits and absorbs light 
at X\,\AA\ when an electron moves between the XY transitions. All else being
equal, a star with lower normalised flux at a rest-frame wavelength of X\,\AA\
indicates that the star has a higher abundance of sodium.

We make use of the detailed chemical abundances calculated by \citep{casey}
from 96,648 APOGEE spectra. This data set includes 15 measured chemical abundances
for all stars (e.g., no missing data), with estimated uncertainties. All chemical
abundances are reported as the logarithmic count of the number of atoms relative
to hydrogen, with a correction to account for the Sun's chemical composition such
that "[Na/H] = 0" indicates a star has the same abundance of sodium relative to
hydrogen as the Sun. \todo{To provide context on the fidelity of the data, in Figure \ref{fig:gce} we show the measured abundance
of [Na/H] with respect to the measured abundance of [Fe/H] for all XX,XXX stars,
where the Sun's abundance is marked.}

The data set also contains other measured properties of the
stars, including their temperature and surface gravity, and whether they are 
likely associated with any gravitationally-bound star cluster. Most stars are
not likely members of any known star cluster; most stars in this dataset are 
described as `field' stars because they have no obvious association.
We do not make use of any of this ancillary data (e.g., temperature, cluster
association) during our search, but we do make use of it to evaluate the performance
of our search.

@@

Costs of generated data sets.
Convergence of generated data sets.
When ther eis a switchover and we are faster.

Expectation that kasarapu and allison should be O(3) and we are O(1) with some overhead.

@@

\section{Appendices}
\subsection{The MML objective function}
\label{app:mml_derivation}
In the following subsections we provide expressions for the message length to
encode a gaussian mixture model, and our prior beliefs on the parameters of that model, before describing our
full objective function.

\subsubsection{The number of mixture components (or groups) $K$}

@@

\subsubsection{The relative weight $w_k$ of each component}

@@

\subsubsection{The mean $\vecmean_k$ and covariance matrix $\veccov_k$ for all
               $K$ Gaussian components}

@@

\subsection{Objective function}

@@

\section{Bounds on components of the message length}

@@

\andy{here}

>>>>

\noindent{}Thus, for any finite mixture of $K$ Gaussian components we have the bounds
%\begin{eqnarray}
%\sum_{k=1}^{K}\log\weight_k & \geq & -K\log{K} \nonumber \\
%\sum_{k=1}^{K}\log\weight_k & \leq & 2K\left(\log{2} - 1\right) - 2\log{2} + \log{N} + \log{\left(N - 2K + 2\right)} \nonumber
%\end{eqnarray}
\begin{equation}
-K\log{K} \quad \geq \quad \sum_{k=1}^{K}\log\weight_k \quad \geq \quad (K - 1)\log{2} - K\log{N} + \log{(N - 2K + 2)}
\end{equation}
\noindent{}which  describe the smallest and largest contribution that the weights can contribute to the message length for any untrialled mixture with $K$ components. When making predictions for future mixtures, we record the optimal weights for each trialled mixture and calculate a \emph{uniformity fraction} $f_u \in [0, 1]$
\begin{equation}
f_u = \frac{\alpha - \sum\log\weight_k}{\alpha + K\log{K}}
\end{equation}
\noindent{}where for brevity $\alpha = (K-1)\log{2} + \log(N-2K+2)-K\log{N}$,
and $f_u$ the uniformity of the mixture from Equation \ref{eq:slw_nonuniform} ($f_u = 0$) to Equation \ref{eq:slw_uniform} ($f_u = 1$). We use estimates of $f_u$ from previously trialled mixtures to make predictions for the uniformity of future,
untrialled mixtures, and the message length required to encode the weights.
Given the theoretical bounds in Equations \ref{eq:slw_uniform} and \ref{ew:slw_nonuniform},
we clip the predictions of the expectation value on the message length required to encode the weights by these bounds.

@@

The negative log-likelihood could be infinitely high given the wrong component parameters.
When predicting the negative log likelihood $-\log{\mathcal{L}(\data|\vectheta)}$ we seek to
predict the \emph{minimum} $-\log{\mathcal{L}(\data|\vectheta)}$ for an untrialled mixture with
$K$ components. We assume that the improvement in $-\log{\mathcal{L}(\data|\vectheta}$ between
subsequent trials of $K$ improves of order $1/K$ \cite{Li|Barron:2000}. We assume a mean model for the minimal message lengths $I$ of previously trialled mixtures
with $K$ components as
\begin{equation}
	-\log{\mathcal{L}(\data|\vectheta)} = \frac{\alpha}{K} + \beta
\end{equation}
\noindent{}where $\alpha$ and $\beta$ are unknowns, and fit these data using a Gaussian
Process \citep{GPs} where the covariance between the data ($K$, $I$) is modelled by a squared exponential kernel 
function \citep{GPs;expsquared} with white noise component added to the diagonal of the covariance matrix. 
We simultaneously optimized the parameters $\alpha$, $\beta$ and the kernel hyperparameters
\cite{BFGS,Geoge}.
Given this Gaussian Process model with the optimized parameters, we make predictions for the minimum message length of untrialled mixtures. Specifically we are making predictions for the expectation value of the minimum message length of untrialled mixtures, and each prediction of the expectation value has an associated variance in its prediction. Given the kernel adopted, the predicted variance will be higher for untrialled mixtures that are farthest away (in $K$) than any trialled mixture (Figure~\ref{fig:mml-predictions}).

@@

The remaining component whose contribution to the message length is unknown is the sum of
the log of the determinant of the covariance matrices $\sum\log|\veccov_k|$. For most
practical scenarios this component is a much smaller order of magnitude than the negative
log likelihood $-\log{\mathcal{L}(\data|\vectheta)}$. Through experiments with generated
data sets, we found that the message length to encode $\sum\log|\veccov_k|$
\begin{equation}
	I_{sldc} = -\frac{1}{2}(D + 2)\sum\log|\veccov_k|
\end{equation}
\noindent{}tended to decrease with increasing $K$ components when the value of $K$ is low
(e.g., $\approx \frac{1}{5}K_{true}$), and then began to increase linearly with $K$ until
$K_{true}$. Our experiments suggested that this is merely a consequence of the search
process, and cannot be reliably used for a heuristic for predicting $K_{true}$, as the
turnover point occurred when the first true data component is well-described by a single
model component. That is to say that by $K \approx \frac{1}{5}K_{true}$, there is one
component covariance matrix that will no longer change significantly, and therefore the
determinants of the component covariance matrices that are added to the mixture will only
tend to decrease on average, producing a near linear increase in $I_{sldc}$. Before the
turnover point, as components are added the average determinants of the covariance matrices
continues to increase.

 % Remove above paragraph?
We trialled several heuristics to approximate the change in $\sum\log|\veccov_k|$ between
mixtures. While some heuristics based on the distribution of distances between data points
appeared promising, we found good results by simply modelling $I_{sldc}$ as a function
of the number of trialled components $K$ using a Gaussian Process. We used a Mat\'ern-3/2
kernel \citep{matern,GPs} function to model the covariance between data points ($K$, $I_{sldc}$) and added
hyperparameters to fit the data mean, and a white noise component that was added to the
diagonal of the covariance matrix. We optimized these hyperparameters using the L-BFGS-B
algorithm \citep{bfgs,scipy,george}, and made predictions for the message length required
to encode the sum of the log of the determinant of covariance matrices of untrialled
mixtures.

Like the predictions from our Gaussian Process model for $-\log{\mathcal{L}(\data|\vectheta)}$,
our predictions for the expectation value of $-\frac{1}{2}(D + 2)\sum\log|\veccov_k|$ (for
the optimal mixture) come with an associated variance on that prediction.

@@

We sum the predictions in the message length required to describe $\sum\log\weight_k$, the
negative log likelihood $-\log{\mathcal{L}(\data|\vectheta)}$, and the sum of the log of 
the determinant of the covariance matrices $\sum\log|\veccov_k|$, as well as constant and
deterministic terms, to predict the (minimum) message length of an untrialled mixture.
We sum the variances in the predictions of each component. This process is illustrated in
Figure~\ref{fig:mml-predictions} where we show the predicted (blue) and actual message lengths 
(black) for various mixtures. The
minimum message length over all models $\vec{\mathcal{M}}$ is predicted to be near
$K \approx \todo{X}$, even though only a few mixtures have been trialled, and the nearest
trialled mixture is $K = \todo{X}$.

@@

We evaluate the message length of each component by rearranging Equation \ref{eq:objective_function} and multiplying the negative log-likelihood
$-\log{\mathcal{L}}(\vec\data|\vectheta)$ by the $N \times K$ responsibility matrix $\mathcal{R}$ 
%\begin{eqnarray}
%	\mathcal{R}_{nk} = \frac{\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}{\sum_{k=1}^{K}\weight_{k}f(\vec\data_n|\vecmean_k,\veccov_k)}
%\end{eqnarray}
such that the message length of the $k$th component is
\begin{eqnarray}
	I_k &=& -\sum_{n=1}^{N}\mathcal{R}_{nk}\log{\weight_{k}f(\data_{n}|\vecmean_k,\veccov_k)} -\frac{(D + 2)}{2}\log{|\cov_k|} \nonumber \\
		&& \cdots\ + \left(\frac{D(D+3)}{4}-\frac{1}{2}\right)\log{\weight_k} + \left(1 - \frac{D}{2}\right)\log{2} \nonumber \\
		&& \cdots + \frac{1}{K}\left[\log\Gamma(K) + \frac{1}{2}\left(\log{Q\pi} - Q\log{2\pi}\right) + \frac{Q}{2}\log{N}\right] \label{eq:component-message-length}
\end{eqnarray}
\noindent{}and the total message length remains the sum of the message lengths of the components
\begin{eqnarray}
	I(\vec\data|\vec\psi, N, D) = \sum_{k=1}^{K}I_{k} \quad .
\end{eqnarray}

@@

When splitting the component with the largest message length, we split this component into two child components. Initially we 
set the means of the child mixtures as positive and negative eigenvalues along the principal
eigenvector of the component covariance matrix \citep{Kasarapu|Allison:2015}.
A matrix of child responsibilities is then
calculated based on the Euclidean distance between each datum and the child means (i.e., hard
association), which we use to estimate the covariance matrices and weights of the child
components. After the parameters of the child components are initialised, we run \emph{partial}
expectation-maximization to iteratively update the parameter estimates of the child components. 
By \emph{partial} expectation-maximization we refer to the process of keeping the parameters of
all other components fixed, and only updating the estimates of the parameters of the child
components \citep{Kasarapu|Allison:2015}. After partial E-M has converged to the same specified tolerance given for the search, we integrate the child components with the rest of the mixture and run full E-M before the \textsc{Split} operation (line \ref{line:mbm-split} of Algorithm~\ref{alg:MBM}) is considered complete.

One advantage of the Message-Breaking Method is that the total number of E-M steps is relatively few because all partial E-M steps are performed from a near-optimal initialisation point. One possible disadvantage to this approach is that there may be mixtures where the \textsc{Split} operation increases the message length by $\epsilon$, but a subsequent \textsc{Split} operation would \emph{decrease} the total message length by $>\epsilon$. As a result, the algorithm would converge to a local minima, not the global one. In the code implementation that we release with this work, we have implemented a \emph{Strict Message-Breaking Method} variant that allows the user to specify the number of successive $K$ trials that increased the message length, before adopting the mixture with the shortest message length. By default the Strict Message-Breaking Method requires five successive mixtures where the message length did not improve.

@@

The \emph{Message-Jumping Method} seeks to find the mixture parameters $\vec\psi$ that produce the minimum message length over the class of finite mixtures $\vec{\mathcal{M}}$ in the fewest number of computations. This objective is shared by the \emph{Message-Breaking Method} and the \emph{Message-Jumping Method}, but they differ in their approach.

In the \emph{Message-Breaking Method} we seek to minimize the total number of computations by using  partial E-M to minimise the total number of (effective) E-M steps
required to converge on a $K+1$ component mixture from a $K$ component mixture.
In the \emph{Message-Jumping Method} we seek to minimise the total number
of computations by optimizing a fewer number of mixtures. Instead of iteratively
perturbing a mixture from say $K = 10$ to $K = 30$, we use predictions of the
message length of untrialled mixtures and immediately `jump' to a component (
e.g., $K = 27$) that will provide the most improvement in our predictions of the
minimum message length over \emph{all} components. In doing so we seek a global 
message length minima with a fewer number of total E-M operations.

We describe the \emph{Message-Jumping Method} in Algorithm~\ref{alg:MBM}. In this
method we seek to predict the minimum message length for any number of $K$ components,
and then converge to the optimal mixture as quickly as possible. For this reason,
we initialise many mixtures that are evenly separated in $\log_{10}$ space from
$K = 1$ to $K = N/2$ (the maximum possible number of components) in order to
quickly describe the minimum message length across all $K$. Specifically we
introduce a search hyperparameter $K_{inits}$ (defaults to 10) that sets the
number of mixtures to initialise -- and those initial $K$ values.

We initialise each mixture using the K-means++ algorithm\cite{kmeanspp} and
optimize the mixture using EM as described earlier. We trial each of the $K_{init}$
mixtures in order of increasing $K$. If one of these mixtures fails to initialise
or optimize, this is likely because of zero entries in the diagonal of the covariance
matrices. In practice this is typically because
the estimate of $K$ is much higher than $K_{true}$. In this situation we do not attempt
to initialise any more of the $K_{init}$ mixtures, and we being the search method.

The motivation for this initialisation process is to mimic the search strategy
that a human might take: What is an order-of-magnitude estimate on the number of
$K$? Or rather: what is a ballpark number on the maximum number of $K$ that we should
trial? After we have that estimate, and we have an estimate on the message length for a
few sparsely-sampled numbers of components, we can make reasonable predictions
on the global minimum.

We record the message length for the optimal mixture at each trialled $K$, and
the model parameters for that mixture. With these quantities we predict the
message length of untrialled mixtures for every $K$ value from $K = 1$ to 
\todo{some number}. The message length predicted from the optimised Gaussian processes
provide an expectation value on the message length for an untrialled mixture, as
well as a variance in that prediction (Figure \ref{fig:search-predictions}).
Before selecting the next value of $K$ to trial, we use the expectation value and
associated variance to compute the acquisition function\cite{Jones:1998} (or the "expected
improvement") at each value of $K$ 
\begin{equation}
	\todo{\Psi = X}
\end{equation}
\noindent{}and select the next $K$ value by choosing the $K$ that will maximize
the expected improvement.

After the next $K$ value has been chosen, we initialise a new mixture with $K$
components using the K-means++ algorithm\citep{kmeanspp} and run E-M until
convergence. If the initialisation and optimization process fails, we reselect
the next $K$ value that will maximise the expected improvement. After the mixture
has converged, we update our predictions for the message length of untrialled
mixtures and choose the next $K$ value to trial.

@@

\caption{The Component-Jumping Method.}
\label{alg:component-jumping-method}
{\bf Input:} $\vec\data$, $K_{i}$, $I_\epsilon$ \\
{\bf Output:} $\vec\psi$ \Comment{The optimal mixture parameters.}
\begin{algorithmic}[1]
\State s $\leftarrow (K_{i})^{-1}\log_{10}{N}$ \Comment{Initial step size in $K$.}
\State v $\leftarrow [0, 1, ..., K_{i}]$ 
\State $K_{inits} \leftarrow$ \Call{Power}{10, vs} \Comment{Set initial $K$ values to trial.}
\State list $K_{t}$ \Comment{Create a list to save trialled $K$ values.}
\For{$k\in K_{inits}$} 
  \State $K_{t}$.append($k$) \Comment{Remember that we trialled this $K$ value.}
  \try 
	  \State mix $\leftarrow$ \Call{KMeans$^{++}$}{k} \Comment{Initialise with K-means++.}
	  \State mix $\leftarrow$ \Call{ExpectationMaximization}{mix}
  \catch
  	  \State $K_{max} \leftarrow k$ \Comment{Record the maximum $K$ value.}
  	  \State {\bf \sc Break} \Comment{Do not trial any higher $K$ values.}
  \finally 
      \State mixtures[k] $\leftarrow$ mix 
      \State states[k] $\leftarrow$ \Call{MessageLength}{mix}
  \endtry
\EndFor
\State converged $\leftarrow$ FALSE
\While{not converged}
	\State $K_{predict} \leftarrow [1, \cdots, K_{max}]$ \Comment{Predict future message lengths.}
	\State $I_{predict} \leftarrow$ \Call{PredictMessageLength}{$K_{predict}$,mixtures,states}
	% argsort.
	\State $A \leftarrow$ \Call{Acquisition}{$K_{predict},I_{predict}$} \Comment{Calculate the expected improvement.}
	\State $K_{next} \leftarrow$ \Call{ArgSort}{$I_{predict}$} \Comment{Sort $K$ by predicted message length.}
	\For{$k\in K_{next}$}
		% Check if the mixture at K was good?
		\If{$k\in K_{t}$}
			\State converged $\leftarrow$ TRUE
			\State {\bf \sc Break}
		\Else
			\State $K_{t}$.append($k$) \Comment{Remember that we trialled this $K$ value.}
	   	  \try 
		    \State mix $\leftarrow$ \Call{KMeans$^{++}$}{k} \Comment{Initialise with K-means++.}
	   	    \State mix $\leftarrow$ \Call{ExpectationMaximization}{mix}
		  \catch
  	        \State {\bf \sc Continue} \Comment{Move on to next $K$ trial.}
	      \finally
	      	\State mixtures[k] $\leftarrow$ mix
			\State states[k] $\leftarrow$ \Call{MessageLength}{mix}
		  \endtry 
		\EndIf
	\EndFor
\EndWhile
\end{algorithmic}
\end{algorithm}

@@

\begin{comment}
We set the data dimensionality $D = 2$ after previous experiments found that the data dimensionality had only a small effect on the differential cost on the search strategies tested. We chose $N$ to vary uniformly in log space from $N = 10$ data points to $N = 10^{4.5}$. For each $N$ we set $K = N/10$, and then used these values of $N$, $D$, and $K$ to generate a synthetic dataset. The weights for each mixture were drawn from a Dirichlet distribution with concentration $\alpha = 100$ while ensuring that $\weight_k \in \left(\frac{2}{N}, 1-\frac{2(K-1)}{N}\right)$. The means for each component were drawn from $\mathcal{U}\left(-10, +10\right)$. The correlation coefficients  for the component covariance matrices were drawn from
\begin{equation}
	\rho \sim \frac{\left[1-\mathcal{U}(-1, +1)^2\right]^{\frac{1}{2}(s - 4)}}{\mathcal{B}\left(\frac{1}{2}, \frac{1}{2}(s - 4)\right)}
\end{equation}
\noindent{}where $s$ is an isotropy hyperparameter that describes the probability of correlations between data dimensions, which we fix as $s=100$. The diagonal variances for the component covariance matrices were drawn from $|\mathcal{N}\left(0, 0.1\right)|$. Given these randomly drawn parameters we generated $N\weight_k$ members for each component, rounding as necessary to ensure the total number of generated data points is kept at $N$.
\end{comment}

@@

We use detailed chemical abundances from the Apache Point Observatory Galactic Evolution Experiment (APOGEE), an experiment run by the Sloan Digital Sky Survey (SDSS) collaboration. We make use of SDSS Data Release 12, which includes high-resolution, high S/N infrared spectra for $\sim$163,000 stars collected by APOGEE. Each observation
is an array of 8,575 pixels that have an associated rest-frame wavelength, and a
measurement of the stellar flux that is normalised relative to the spectral energy
distribution such that a flux value of 1 indicates no stellar absorption and a
flux value near 0 implies strong stellar absorption. All else being equal, a stronger
amount of absorption (lower normalised flux value) at a particular wavelength 
indicates that the star contains more of the chemical element that absorbs and
emits light at that wavelength. For example, sodium emits and absorbs light 
at 1639.33\,nm in vacuum conditions. All else being
equal, a star with lower normalised flux at a rest-frame wavelength of 1639.33\,nm
indicates that the star has a higher abundance of sodium.

We make use of detailed chemical abundances measured for 96,648 APOGEE spectra\citep{Casey:2016}.
This data set includes 15 measured chemical abundances
for all stars (e.g., no missing data), with estimated uncertainties. All chemical
abundances are reported as the logarithmic count of the number of atoms relative
to hydrogen, with a correction to account for the Sun's chemical composition such
that a value of [Na/H] = 0 indicates a star has the same abundance of sodium relative to
hydrogen as the Sun. To provide some context for the data, in Figure \ref{fig:gce} 
we show the measured abundance of various elements relative to the measured abundance
of [Fe/H] for all 96,648 stars. The Sun's abundance is marked.

The data set also contains other measured properties for each star. This 
includes the effective temperature and surface gravity, and whether they are 
likely associated with any gravitationally-bound star cluster. Most stars are
not likely members of any known star cluster; most stars in this dataset are 
described as `field' stars because they have no obvious association.
We do not make use of any of this ancillary data (e.g., temperature, cluster
association) during our search, but we do make use of it after the search is
complete in order to evaluate the performance of our search.

@@

We recorded the wall clock time required by each search algorithm on generated data sets, which we show in Figure~\ref{fig:cost}. Although wall-clock time metrics are subject to decisions made during implementation, all search and optimisation methods are implemented in the same way (using the same objective function), so the relative differences between methods is reliable. The \emph{Message-Breaking Method} and \emph{Message-Jumping Method} have approximately \todo{the same costs in wall time as a function of the number of clusters}.

\todo{Costs of generated data sets.}
\todo{Switchover in time and we are faster}

\todo{Convergence of generated data sets.}

\todo{Expected complexity of \cite{Kasarapu|Allison:2015} given the search setup}

The \emph{Message-Breaking Method} splits the component with the largest message length (Equation~\ref{eq:component-message-length}), whereas the benchmark method~\cite{Kasarapu|Allison:2015} tries to split, merge, and delete each component. After repeated trials on generated data sets, we found that the benchmark method~\cite{Kasarapu|Allison:2015} would most frequently chose to split the same component that the \emph{Message-Breaking Method} would split (e.g., the component with the largest message length). We tracked the operation type (e.g., split, delete, merge) and which component was operated upon for \todo{X} operations in \todo{Y} generated data sets for the benchmark method~\cite{Kasarapu|Allison:2015} and the \emph{Message-Breaking Method}.

The benchmark method chose to split in all \todo{X} operations (e.g., no delete or merge operations). We also found that the benchmark method usually (X/X; X\%) chose to split the same component that the \emph{Message-Breaking Method} split: the component with the largest message length. The results are shown in Figure~\ref{fig:path-operations}. Some X\% of the time the component with the second largest message length was split by the benchmark method, X\% of the time with the third-largest, and X\% with the fourth-largest. Despite there being X operations with more than 4 components available to split, the benchmark method consistently chose one of four components with the  largest message lengths. Despite the path differences taken between the benchmark method and the \emph{Message-Breaking Method}, the optimised mixture parameters and message lengths found by each search algorithm were usually the same within the E-M convergence tolerance. This indicates that the \emph{Message-Breaking Method} is just as effective as the benchmark method on generated data sets, and more efficient because it does not need to trial all possible operations on all possible components.

%One advantage of greedily trying to operate on all components is that there is confidence that `the best' operation has been taken, as all other possible operations have been trialled. Cost is the most obvious disadvantage to such an approach.

@@

\appendix

\section{The MML objective function}
\label{app:mml_derivation}
In the following subsections we provide expressions for the message length to
encode a gaussian mixture model, and our prior beliefs on the parameters of that model, before describing our
full objective function.

\subsection{The number of mixture components (or groups) $K$}

@@

\subsection{The relative weight $w_k$ of each component}

@@

\subsection{The mean $\vecmean_k$ and covariance matrix $\veccov_k$ for all
               $K$ Gaussian components}

@@

\subsection{The complete objective function}

@@

\begin{comment}
\section{Bounds on components of the message length}

@@

\andy{here}
\end{comment}
